---
title: "spring migration phenology"
author: "Ellen Blomme"
date: "2024-04-18"
output: html_document
---
# IMPORT AND CLEAN THE DATA

```{r setup, include=FALSE}
# visualization
library(tidyverse)
library(hrbrthemes)
library(viridis)

# setup
library(readxl)
library(janitor)

#model
library(lme4)
library(lmerTest)
library(mgcv)
library(car)

# model selection
library(lmtest)

# spatial
library(sf)
library(sp)
library(gstat)
library(ggspatial)

# vizualize the trends
library(gratia)

# check for correlation
library(ggcorrplot)
library(corrplot)
```


## IMPORT AND CLEANING DATA
```{r import data & cleaning}
amfib <- read_delim("PaddenoverzetDag_2023_01_25 12_26_50.csv", delim = ";") %>% 
  clean_names() %>%
  mutate(line = row_number()) %>%
  filter(jaar != 2023)

sites <- amfib %>% 
  distinct(gemeente, naam) %>%
  mutate(site_code = paste0(naam, ", ", gemeente))

amfib <- amfib %>% 
  mutate(naam = toupper(gsub("[*]","", naam))) %>%
  mutate(datum = gsub("1/01/1970", NA, datum)) %>% # There were no toad patrols in 1970
  mutate_if(is.logical, as.numeric)

sites <- amfib %>% 
  distinct(provincie, gemeente, naam) %>%
  mutate(gemeente = gsub("MEEUWEN-GRUITRODE", "MEEUWEN", gemeente)) %>%
  mutate(gemeente = gsub("GLABBEEK-ZUURBEMDE", "GLABBEEK", gemeente)) %>%
  mutate(site_code = paste0(naam, ", ", gemeente)) %>%
  arrange(site_code) %>%
  mutate(site_id = row_number()) %>%
  dplyr::select(site_id, site_code, naam, gemeente, provincie)

amfib <- amfib %>% 
  mutate(site_code = paste0(naam, ", ", gemeente)) %>%
  left_join(dplyr::select(sites, site_id, site_code), by="site_code")

amfib <- amfib %>% 
  mutate(jaar = as.numeric(jaar)) %>%
  mutate(datum = as.Date(datum, "%d/%m/%Y"))

categorie <- read_csv("categorie.csv") %>% 
  dplyr::select(-c(n))

amfib_long <- amfib %>%
  dplyr::select(site_id, jaar, datum:salamanders_kolken) %>%
  pivot_longer(!c(site_id:temperatuur), names_to = "categorie", values_to = "count")

amfib_long <- amfib_long %>% left_join(categorie, by="categorie")

site_jaar_overview_dagen <- amfib_long %>% 
  filter(!is.na(datum) & !is.na(jaar) & !is.na(site_id)) %>%
  distinct(site_id, jaar, datum) %>%
  group_by(site_id, jaar) %>%
  dplyr::summarise(tot_jaar_dagen = n()) %>%
  ungroup()

sites_summary <- sites %>%
  left_join(site_jaar_overview_dagen %>%
              group_by(site_id) %>% 
              dplyr::summarise(jaren_alle = n()), by="site_id")
```

```{r set up the proper dataset}
# only select the toad counts without any missing values for the toad count
# REMARK - 1 observation has 2012 for year but 1/1/1970 for date
bufo_bufo <- amfib_long[amfib_long$soort == "PAD" & 
                          !is.na(amfib_long$count) & !is.na(amfib_long$datum), ]

# join site info with count data
bufo_bufo <- merge(x = bufo_bufo, y = sites, 
                   by = "site_id", all.x = TRUE)

# Molsbroek == Waasmunsterbaan
bufo_bufo$site_id <- ifelse(bufo_bufo$site_id == 364, 585, bufo_bufo$site_id)
bufo_bufo$site_code <- ifelse(bufo_bufo$site_code == "MOLSBROEK, LOKEREN", "WAASMUNSTERBAAN, LOKEREN", bufo_bufo$site_code)

# there is only one count for a couple
bufo_bufo$count <- ifelse(bufo_bufo$geslacht == "K", bufo_bufo$count * 2, bufo_bufo$count) 

# change the year of december to the following year so it is correct to calculate
# the shift in phenology
bufo_bufo$jaar <- ifelse(format(bufo_bufo$datum, "%m") == 12, 
                          bufo_bufo$jaar + 1, 
                          bufo_bufo$jaar)

# Calculate the cumulative count for a date per location per year
bufo_bufo <- bufo_bufo %>%
  # remove all counts of toads going back to the forest/terrestrial habitat
  filter(categorie != "gew_pad_terug") %>% 
  arrange(site_id, jaar, datum) %>% 
  group_by(site_id, jaar) %>%
  # total amount of migrated toads migrated on that day (per location per year)
  mutate(cumulative_count = cumsum(count)) %>% 
  ungroup()
```

```{r only look at locations with more than 5 years of data}
# remove sites with < 5 years of counts
bufo_bufo <- filter(bufo_bufo, site_id %in% 
                 sites_summary$site_id[sites_summary$jaren_alle > 5])
sites_summary <- filter(sites_summary, site_id %in% 
                          sites_summary$site_id[sites_summary$jaren_alle > 5])
```


# EXPLORATION

# DISTRIBUTION OF SAMPLING POINTS
```{r read in spatial data}
gemeentes <- st_read("Gemeentes_Vlaanderen.shp") %>%
  rownames_to_column("FID") %>% dplyr::select(-FID_) # GIS data

# clean the data
acties <- read_xlsx("PaddenoverzetActies_2023_01_25 12_26_27.xlsx") %>% 
  clean_names() %>%
  mutate(line = row_number()) %>%
  mutate(naam = toupper(gsub("[*]","", naam))) %>%
  mutate(gemeente = gsub("MEEUWEN-GRUITRODE", "MEEUWEN", gemeente)) %>%
  mutate(gemeente = gsub("GLABBEEK-ZUURBEMDE", "GLABBEEK", gemeente)) %>%
  mutate(site_code = paste0(naam, ", ", gemeente)) # all toad actions
# filter(!(latitude == 0 | is.na(latitude)) &
# !(longitude == 0 | is.na(longitude))) # remove all incomplete coordinates 
# use, to find the empty coordinates


# fulfill the coordinates of locations
acties[acties$site_code == "BAVERSTRAAT (ELDEREN), TONGEREN",
       "latitude"] <- 50.793217874202576
acties[acties$site_code == "BAVERSTRAAT (ELDEREN), TONGEREN",
       "longitude"] <- 5.487557081429292

acties[acties$site_code == "BORGGRAVEVIJVERSTRAAT, HASSELT",
       "latitude"] <- 50.94910703887022
acties[acties$site_code == "BORGGRAVEVIJVERSTRAAT, HASSELT",
       "longitude"] <- 5.3737536814389975

acties[acties$site_code == "BROEKSTRAAT, FONTEINSTRAAT, WILLEBROEK",
       "latitude"] <- 51.059856735134396
acties[acties$site_code == "BROEKSTRAAT, FONTEINSTRAAT, WILLEBROEK",
       "longitude"] <- 4.3811436967902

acties[acties$site_code == "OUDE GENTWEG, AALTER",
       "latitude"] <- 51.07913415602838
acties[acties$site_code == "OUDE GENTWEG, AALTER",
       "longitude"] <- 3.477390025627469

acties[acties$site_code == "SULFERBERG (WESTOUTER), HEUVELLAND",
       "latitude"] <- 50.79674300378389
acties[acties$site_code == "SULFERBERG (WESTOUTER), HEUVELLAND",
       "longitude"] <- 2.767450954685736
```

```{r make a map}
site_coords.sf <- acties %>%
  mutate_at(c("latitude", "longitude"), ~na_if(., 0)) %>%
  group_by(site_code) %>%
  summarise(lat = mean(latitude, nan.rm = T),
            lon = mean(longitude, nan.rm = T))%>%
  filter(!is.na(lat)) %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant")

# sample sizes and longest time series
site.sf <- site_coords.sf %>% 
  right_join(sites_summary, by="site_code")

bbox_VL <- st_bbox(gemeentes %>% st_transform(crs=4326))

#making a map of Flanders with the locations
main_map <- ggplot() +
  theme_bw() + 
  geom_sf(data = st_geometry(gemeentes %>% st_transform(crs=4326)), colour = "grey", fill=NA, lwd = .5) +
  geom_sf(data = site.sf, shape=19, size = 1, alpha=8)+
  scale_x_continuous(limits = c(2.66, 5.8), position="top") +
  scale_y_continuous(limits = c(50.15, 51.48)) +
  annotation_scale(location="tl", width_hint = 0.6,
                   pad_y = unit(6.9, "cm")) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.1, "cm"), pad_y = unit(0.4, "cm"),
                         style = north_arrow_fancy_orienteering)

main_map


#making a map of NW-Europe for situating Flanders


world <- ne_countries(scale="large", returnclass="sf")
target_crs <- 4326#"+proj=longlat" #+proj=longlat CRS("EPSG:4236)

world_wgs84 <- world %>% st_transform(crs = target_crs)

inset_map <- world_wgs84 %>% ggplot() + geom_sf() + #map of europe
  scale_x_continuous(limits = c(-5, 15)) +
  scale_y_continuous(limits = c(48, 55)) +
  geom_rect(data = data.frame(),
            aes(xmin = bbox_VL$xmin, xmax = bbox_VL$xmax, ymin = bbox_VL$ymin, ymax = bbox_VL$ymax),
            colour = "black", fill=NA, size=1)+
  theme_bw()+
  theme(plot.background = element_rect(colour = "black", fill=NA, linewidth=0.5))

inset_map



ggdraw(main_map) +
  draw_plot(
    {
      inset_map
    },
    # The distance along a (0,1) x-axis to draw the left edge of the plot
    x = 0.605, 
    # The distance along a (0,1) y-axis to draw the bottom edge of the plot
    y = 0.12,
    # The width and height of the plot expressed as proportion of the entire ggdraw object
    width = 0.35, 
    height = 0.35)

# ggsave("figure 1.jpeg", width = 13.9, height = 6,  units = "cm", dpi=600)
```


# CALCULATE BEGIN, PEAK, END

```{r median 5th and 95th percentile}
bufo <- bufo_bufo %>%
  dplyr::rename(year = jaar) %>%
  group_by(year, site_id, datum) %>%
  dplyr::summarise(count = sum(count), # sum of dead and migrating toads
                   migr_toads = max(cumulative_count), # total migrated toads on that day
                   ) 

# Get the estimates for the median, 5th and 95th percentile per id and year
estimate <- bufo %>%
  group_by(year, site_id) %>%
  dplyr::summarise(median_count = median(migr_toads),
                   first_perc = quantile(migr_toads, 0.05), 
                   #quantile(count, 0.05)
                   last_perc = quantile(migr_toads, 0.95),
                   tot_count = sum(count),
                   tot_days = n_distinct(datum)) %>%
  ungroup()

# Merge the datasets based on year and location
merged_bufo <- merge(bufo_bufo, estimate, by.x=c("site_id", "jaar"), 
                     by.y=c("site_id", "year"))

# remove columns that are not useful
merged_bufo <- subset(merged_bufo, select = -c(tijd, regen, soort,
                                               beweging, geslacht, naam, 
                                               gemeente, provincie))

# remove columns with missing id
merged_bufo <- merged_bufo[!is.na(merged_bufo$site_id), ]
```



```{r min date per statistic}
# find the date for the peak of the migration
peak_data <- merged_bufo %>%
  dplyr::rename(year = jaar) %>%
  group_by(site_id, year) %>%
  # select the first day where the amount of toads crossed is higher than the
  # median count
  filter(cumulative_count >= median_count) %>%
  # The day of the year as a decimal number (range 001 to 366).
  summarise(peak = as.numeric(format(min(datum), format = "%j"))) 

# find the date for the start of the migration, similar to the peak
begin_data <- merged_bufo %>%
  dplyr::rename(year = jaar) %>%
  group_by(site_id, year) %>%
  filter(cumulative_count >= first_perc) %>%
  summarise(onset = as.numeric(format(min(datum), format = "%j")))

# find the date for the end of the migration, similar to the peak
end_data <- merged_bufo %>%
  dplyr::rename(year = jaar) %>%
  group_by(site_id, year) %>%
  filter(cumulative_count >= last_perc) %>%
  summarise(end = as.numeric(format(min(datum), format = "%j")))
```

```{r joined dataset (amount of rows)}
phenology <- begin_data %>%
  left_join(peak_data, by = c("site_id", "year")) %>%
  left_join(end_data, by = c("site_id", "year")) %>%
  left_join(estimate[, c("site_id", "year", 
                         "tot_count", "tot_days")], 
            by = c("site_id", "year")) %>%
  mutate(year_center = year - 2020,
         site_id = as.factor(site_id),
         migration_length = end - onset + 1) %>% # filter(tot_days > 1)
  filter(migration_length > 1)
```




# ANALYSIS MIGRATION PHENOLOGY
Models testing whether the beginning / peak / end of breeding migrations shifted
over the years. As the migration may shift depending on the location, we use a 
random intercept for each location.

## MIXED MODELS

### ONSET
```{r model of STARTdate}
mm_onset <- lmer(onset ~ year_center + (1|site_id),
              data = phenology)

complex <- lmer(onset ~ year_center + (1+year_center|site_id),
              data = phenology) # is the change consistent

summary(mm_onset)
```

```{r model selection of STARTdate}
# Are the populations reacting in the same way?
#H0: The full model and the nested model fit the data equally well.
#Ha: The full model fits the data significantly better than the nested model.
lrtest(mm_onset, complex)
```

```{r onset assumptions}
plot(phenology$year_center, phenology$onset)
lines(smooth.spline(phenology$onset~phenology$year_center))

# Residual Plot
residuals <- resid(mm_onset)
fitted_values <- fitted(mm_onset)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```


### PEAK
```{r model of PEAKdate}
mm_peak <- lmer(peak ~ year_center + (1|site_id),
              data = phenology)
complex_peak <- lmer(peak ~ year_center + (1+year_center|site_id),
              data = phenology)

summary(mm_peak)
```

```{r model selection of PEAK}
# Are the populations reacting in the same way
#H0: You should use the complex model.
#Ha: You should use the nested model.
lrtest(mm_peak, complex_peak)
```

```{r peak assumptions}
plot(phenology$year_center, phenology$peak)
lines(smooth.spline(phenology$peak~phenology$year_center))

# Residual Plot
residuals <- resid(mm_peak)
fitted_values <- fitted(mm_peak)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```


### END
```{r model of ENDdate}
mm_end <- lmer(end ~ year_center + (1|site_id),
              data = phenology)
complex_end <- lmer(end ~ year_center + (1+year_center|site_id),
              data = phenology)

summary(mm_end)
sum_end <- summary(complex_end)
sum(sum_end$coefficients[, "Pr(>|t|)"] <= 0.05)
```

```{r model selection of END}
# Are the populations reacting in the same way
#H0: You should use the complex model.
#Ha: You should use the nested model.
lrtest(mm_end, complex_end)
```


```{r end assumptions}
plot(phenology$year_center, phenology$end)
lines(smooth.spline(phenology$end~phenology$year_center))

# Residual Plot
residuals <- resid(mm_end)
fitted_values <- fitted(mm_end)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```



### DURATION
```{r model of migration DURATION}
mm_length <- lmer(migration_length ~ year_center + (1|site_id),
                  data = phenology)
complex_length <- lmer(migration_length ~ year_center + (1+year_center|site_id),
                       data = phenology)
summary(mm_length)
```

```{r model selection of DURATION}
# Are the populations reacting in the same way?
#H0: You should use the complex model.
#Ha: You should use the nested model.
lrtest(mm_length, complex_length)
```

```{r is there variation over the years in duration}
gamm_int <- gam(migration_length ~ s(year_center) + s(site_id, bs = "re") + s(site_id,
    year_center, bs = "re"), data = phenology, method = "REML")
lmm_int <- gam(migration_length ~ year_center + s(site_id, bs = "re") + s(site_id,
    year_center, bs = "re"), data = phenology, method = "REML")

# Is the duration variable?
#H0: You should use the complex model.
#Ha: You should use the nested model.
lrtest(lmm_int, gamm_int)
```

```{r duration assumptions}
plot(phenology$year_center, phenology$migration_length)
lines(smooth.spline(phenology$migration_length~phenology$year_center))

# Residual Plot
residuals <- resid(mm_length)
fitted_values <- fitted(mm_length)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```

# MODEL VISUALISATION
```{r plots dates}
phenology$pred_start <- predict(mm_onset, newdata = phenology, re.form = NA)
phenology$pred_end <- predict(mm_end, newdata = phenology, re.form = NA)
phenology$pred_peak <- predict(mm_peak, newdata = phenology, re.form = NA)

plot_pheno <- phenology %>%
  pivot_longer(cols = onset:end, 
               names_to = "date", 
               values_to = "day_of_year") 
plot_pheno$date <- factor(plot_pheno$date, levels = c("onset", "peak", "end"))

ggplot(plot_pheno, aes(x = year, y = day_of_year, color = date))  +  
  theme_classic() +
  theme(legend.position = c(0.1, 0.23)) +
  scale_color_manual(values = c("onset" = "darkgreen", 
                                "peak" = "deepskyblue2", 
                                "end" = "firebrick")) +
  geom_point(size = 1) +
  geom_line(aes(y = pred_start), 
            color = "darkgreen", size = 1) +
  geom_line(aes(y = pred_peak), 
            color = "deepskyblue2", size = 1) +
  geom_line(aes(y = pred_end), 
            color = "firebrick", size = 1) +
  labs(x = "Year", y = "Migration date (day of the year)")


# ggsave("figure 2.jpeg", width = 13.9, height = 8,  units = "cm", dpi=600)
```


# SPATIAL CORRELATION
check if we need to take into account any spatial correlation.
```{r combine with spatial data}
# combine dataset
sites_loc <- sites_summary %>%
  left_join(acties[, c("site_code", "longitude", "latitude")], by = "site_code") %>%
  filter(!(latitude == 0 | is.na(latitude)) &
           !(longitude == 0 | is.na(longitude))) %>%
  distinct(site_id, .keep_all = TRUE)

sites_loc$site_id <- as.factor(sites_loc$site_id)
```

```{r from wgs84 to epsg31370}
sites_loc <- st_as_sf(sites_loc, coords = c("longitude", "latitude"), 
                          crs = 4326, agr = "constant") # WGS84 = EPSG4326
sf_data <- st_transform(sites_loc, crs = 31370) # Belgian Lambert 72

coord <- st_coordinates(sf_data)

sf_data <- cbind(sf_data, coord)

colnames(sf_data)[8] <- "latitude"
colnames(sf_data)[7] <- "longitude"
```


```{r spatial correlation}
# combine dataset
spat <- phenology %>%
  left_join(sf_data[, c("site_id", "longitude", "latitude")], by = "site_id")
phenology <- phenology %>%
  left_join(sf_data[, c("site_id", "longitude", "latitude")], by = "site_id")

# Create a spatial object
coordinates(spat) <- c("longitude", "latitude") # needs to be in Lambert 1972

plot(variogram(spat$peak ~ 1, locations = coordinates(spat), data = spat, cloud = F), 
    type = "b")
```
Based on the correlogram, there is no spatial correlation.


# RELATION WITH TEMPERATURE

```{r read in temperature data and merge}
temperature <- read_xlsx("temperature.xlsx")

# adapt to the right structure
temperature$TEMP_CH_0H_MIN <- as.numeric(temperature$TEMP_CH_0H_MIN)
temperature <- na.omit(temperature)
temperature$TEMP_CH_0H_MAX <- as.numeric(temperature$TEMP_CH_0H_MAX)
temperature$TEMP_CH_0H_AVG <- as.numeric(temperature$TEMP_CH_0H_AVG)
temperature$TIMESTAMP <- as.Date(temperature$TIMESTAMP, format = "%Y-%m-%d")

# Extract day of the year
temperature$day_of_year <- as.integer(yday(temperature$TIMESTAMP))

# Extract year
temperature$year <- as.integer(year(temperature$TIMESTAMP))

# add temperature meterological winter
## Make sure the winter is included in the same toad season
temperature$year <- ifelse(format(temperature$TIMESTAMP, "%m") == 12, 
                          temperature$year + 1, 
                          temperature$year)
temperature$day_of_year <- ifelse(format(temperature$TIMESTAMP, "%m") == 12, 
                          temperature$day_of_year - 365, 
                          temperature$day_of_year)
## select only winter months
meteorological_winter <- subset(temperature, 
                                format(TIMESTAMP, "%m") %in% c("12", "01", "02"))

# Calculate the mean temperature for each year
meteorological_winter <- aggregate(cbind(TEMP_CH_0H_MIN, TEMP_CH_0H_AVG, TEMP_CH_0H_MAX) 
                       ~ year, data = meteorological_winter, FUN = mean)

names(meteorological_winter) <- c("year", "winter_min", 
                                  "winter_avg", "winter_max")
```

```{r ONSETdata}
onset_temp <- phenology %>%
  left_join(meteorological_winter, by = "year") %>%
  left_join(temperature, by = c("year" = "year", "onset" = "day_of_year")) %>%
  dplyr::rename(T_min = TEMP_CH_0H_MIN,
                T_avg = TEMP_CH_0H_AVG, 
                T_max = TEMP_CH_0H_MAX)

onset_temp$mean_week <-  rep(NA, nrow(onset_temp))
onset_temp$mean_month <-  rep(NA, nrow(onset_temp))

for (index in 1:nrow(onset_temp)) {
  # select the data from the dataset
  observation <- onset_temp[index, ]
  
  # select the year and the date
  jaar <- observation$year
  dag <- observation$onset
  
  # calculate the mean avg temperature a week/month before the onset
  weekly_T <- subset(temperature[temperature$year == jaar,], 
                     day_of_year >= dag - 7 & day_of_year <= dag)
  monthly_T <- subset(temperature[temperature$year == jaar,],
                      day_of_year >= dag - 30 & day_of_year <= dag)
  
  # save mean temperature
  onset_temp[index, ]$mean_week <- mean(weekly_T$TEMP_CH_0H_AVG)
  onset_temp[index, ]$mean_month <- mean(monthly_T$TEMP_CH_0H_AVG)
}
```



```{r PEAKdata}
peak_temp <- phenology %>%
  left_join(meteorological_winter, by = "year") %>%
  left_join(temperature, by = c("year" = "year", "peak" = "day_of_year")) %>%
  dplyr::rename(T_min = TEMP_CH_0H_MIN,
                T_avg = TEMP_CH_0H_AVG, 
                T_max = TEMP_CH_0H_MAX)

peak_temp$mean_week <-  rep(NA, nrow(peak_temp))
peak_temp$mean_month <-  rep(NA, nrow(peak_temp))

for (index in 1:nrow(peak_temp)) {
  # select the data from the dataset
  observation <- peak_temp[index, ]
  
  # select the year and the date
  jaar <- observation$year
  dag <- observation$peak
  
  weekly_T <- subset(temperature[temperature$year == jaar,], 
                     day_of_year >= dag - 7 & day_of_year <= dag)
  monthly_T <- subset(temperature[temperature$year == jaar,],
                      day_of_year >= dag - 30 & day_of_year <= dag)
  
  peak_temp[index, ]$mean_week <- mean(weekly_T$TEMP_CH_0H_AVG)
  peak_temp[index, ]$mean_month <- mean(monthly_T$TEMP_CH_0H_AVG)
}
```

```{r ENDdata}
end_temp <- phenology %>%
  left_join(meteorological_winter, by = "year") %>%
  left_join(temperature, by = c("year" = "year", "end" = "day_of_year")) %>%
  dplyr::rename(T_min = TEMP_CH_0H_MIN,
                T_avg = TEMP_CH_0H_AVG, 
                T_max = TEMP_CH_0H_MAX)

end_temp$mean_week <-  rep(NA, nrow(end_temp))
end_temp$mean_month <-  rep(NA, nrow(end_temp))

for (index in 1:nrow(end_temp)) {
  # select the data from the dataset
  observation <- end_temp[index, ]
  
  # select the year and the date
  jaar <- observation$year
  dag <- observation$end
  
  # calculate the mean avg temperature a week/month before the onset
  weekly_T <- subset(temperature[temperature$year == jaar,], 
                     day_of_year >= dag - 7 & day_of_year <= dag)
  monthly_T <- subset(temperature[temperature$year == jaar,],
                      day_of_year >= dag - 30 & day_of_year <= dag)
  
  # save mean temperature
  end_temp[index, ]$mean_week <- mean(weekly_T$TEMP_CH_0H_AVG)
  end_temp[index, ]$mean_month <- mean(monthly_T$TEMP_CH_0H_AVG)
}
```


## MODEL
### ONSET
```{r effect of temperature on ONSET}
onset_week <- lmer(onset ~ mean_week + (1|site_id),
                 data = onset_temp)
summary(onset_week)

onset_month <- lmer(onset ~ mean_month + (1|site_id),
                 data = onset_temp)
summary(onset_month)

onset_winter <- lmer(onset ~ winter_avg + (1|site_id),
                 data = onset_temp)
summary(onset_winter)
```

```{r Tonset assumptions}
plot(onset_temp$winter_avg, onset_temp$onset)
lines(smooth.spline(onset_temp$onset~onset_temp$winter_avg))

# Residual Plot
residuals <- resid(onset_winter)
fitted_values <- fitted(onset_winter)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 32)
shapiro.test(residuals)
```


### PEAK
```{r effect of temperature on PEAK}
peak_week <- lmer(peak ~ mean_week + (1|site_id),
                 data = peak_temp)
summary(peak_week)

peak_month <- lmer(peak ~ mean_month + (1|site_id),
                 data = peak_temp)
summary(peak_month)

peak_winter <- lmer(peak ~ winter_avg + (1|site_id),
                 data = peak_temp)
summary(peak_winter)
```


```{r Tpeak assumptions}
plot(peak_temp$winter_avg, peak_temp$peak)
lines(smooth.spline(peak_temp$peak ~ peak_temp$winter_avg))

# Residual Plot
residuals <- resid(peak_winter)
fitted_values <- fitted(peak_winter)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 32)
shapiro.test(residuals)
```



### END
```{r effect of temperature on END}
end_week <- lmer(end ~ mean_week + (1|site_id),
                 data = end_temp)
summary(end_week)

end_month <- lmer(end ~ mean_month + (1|site_id),
                 data = end_temp)
summary(end_month)

end_winter <- lmer(end ~ winter_avg + (1|site_id),
                 data = end_temp)
summary(end_winter)
```

```{r Tend assumptions}
plot(end_temp$winter_avg, end_temp$end)
lines(smooth.spline(end_temp$end ~ end_temp$winter_avg))

# Residual Plot
residuals <- resid(end_winter)
fitted_values <- fitted(end_winter)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```


### DURATION
```{r effect of temperature on DURATION}
duration_winter <- lmer(migration_length ~ winter_avg + (1|site_id),
                 data = end_temp)
summary(duration_winter)
```

```{r Tduration assumptions}
plot(end_temp$winter_avg, end_temp$migration_length)
lines(smooth.spline(end_temp$migration_length ~ end_temp$winter_avg))

# Residual Plot
residuals <- resid(duration_winter)
fitted_values <- fitted(duration_winter)

# Plot residuals against fitted values
plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(h=0)

# Normality Check
# Q-Q Plot
qqPlot(residuals)
hist(residuals, breaks = 60)
shapiro.test(residuals)
```

## TEMPERATURE VISUALIZATIONS
```{r plots temperatures}
onset_temp$pred_week <- predict(onset_week, newdata = onset_temp, re.form = NA)
onset_temp$pred_month <- predict(onset_month, newdata = onset_temp, re.form = NA)
onset_temp$pred_winter <- predict(onset_winter, newdata = onset_temp, re.form = NA)

plot_temp <- onset_temp %>%
  pivot_longer(cols = c(winter_avg, mean_month, mean_week), 
               names_to = "period", 
               values_to = "temperature") %>%
  select(site_id, onset, period, temperature, pred_week, pred_month, pred_winter)
plot_temp$period <- factor(plot_temp$period, levels = c("winter_avg", "mean_month", "mean_week"))

ggplot(plot_temp, aes(x = temperature, y = onset, color = period))  +  
  theme_classic() +
  theme(legend.position = c(0.123, 0.22)) +
  scale_color_manual(values = c("mean_week" = "darkgreen", 
                                "winter_avg" = "deepskyblue2", 
                                "mean_month" = "firebrick")) +
  geom_point(size = 1) +
  geom_line(data = plot_temp[plot_temp$period == "mean_week",], aes(y = pred_week), 
            color = "darkgreen", size = 1) +
  geom_line(data = plot_temp[plot_temp$period == "mean_month",], aes(y = pred_month), 
            color = "firebrick", size = 1) +
  geom_line(data = plot_temp[plot_temp$period == "winter_avg",], aes(y = pred_winter), 
            color = "deepskyblue2", size = 1) + 
  labs(x = "Temperature (°C)", y = "Onset migration date (day of the year)")
  #, title = "Shift in migration phenology of the common toad (Bufo bufo)"


ggsave("figure 3.jpeg", width = 13.9, height = 8,  units = "cm", dpi=600)
```
